Hydrous melting and partitioning in and above the mantle transition zone: insights from water-rich MgO-SiO$_2$-H$_2$O experiments

### 2016-05-11 revised_submission/MSH.tex
update

%% Group authors per affiliation:
\author{R. Myhill}
\address{Bayerisches Geoinstitut, Universit\"{a}t Bayreuth, Universit\"{a}tsstrasse 30, 95447 Bayreuth, Germany \\ now at School of Earth Sciences, University of Bristol, Wills Memorial Building, Queens Road, Bristol BS8 1RJ, United Kingdom}
\cortext[mycorrespondingauthor]{Corresponding author: R. Myhill}
\ead{myhill.bob@gmail.com}

@@

\section{Introduction}
Hydrous melts have a remarkable influence on Earth's surface and interior. Melting at subduction zones is primarily driven by the release of hydrous fluids from the warming slab at about 100 km depth, producing the world's arc volcanoes. The water cycle does not stop here, however; hydrous minerals within the crust should allow some water to be transported to 250 km \citep{PS2002}, and peridotite hydration should allow water to penetrate into the mantle transition zone or deeper, providing it remains low enough in temperature \citep{KOM2005}. This deep water cycle must extend to at least 500 km depth, as implied by the discovery of water-saturated ringwoodite in a diamond from Juina, Brazil \citep{Pearsonetal2014}. In addition to kimberlites and arc volcanoes, hydrous melts are known to metasomatise and refertilise the lithospheric mantle, creating the so-called MARID (mica, amphibole, rutile, ilmenite, diopside) assemblage. It has also been argued that neutral or negatively buoyant water-rich melts are responsible for low velocity and high conductivity layers observed at the 410 km discontinuity, and that these small fraction melts could act to filter out incompatible elements during upwelling \citep{BK2003}.

@@

Melt compositions are also required to understand the stability of dense hydrous magnesium silicate (DHMS) phases and the water capacity of nominally anhydrous minerals (NAMS) at high pressure. Together, these control water transport and storage in the deep mantle, and as such are the primary inputs to models of the water budget of the Earth \citep{Hirschmann2006}. Hydrous fluids at elevated pressures and temperatures cannot be treated as pure H$_2$O. As pressure increases, immiscibility between fluid and melt decreases \citep{XXXX}, and the thermal stability of hydrous phases increases \citep{XXXX}. As a result, fluids/melts released by the breakdown of hydrous phases have H$_2$O activities significantly lower than one. Understanding and predicting the \emph{P-T} conditions of DHMS phase breakdown and the water storage capacity at of NAMS therefore requires good activity-composition models for hydrous melts.

Constraining melt compositions remains a major challenge in high-pressure experimental petrology. The primary problem is that water-rich melts are unquenchable. During quenching, some of the water in the melt crystallises as hydrous minerals such as brucite, but a significant proportion crystallises as water ice at high pressure, which is then lost as water vapour on decompression and during sample preparation. The original H$_2$O contents of the partial hydrous melts are normally estimated by mass balance and/or by deficits in microprobe totals \citep[e.g.][]{OMY2000, DDFK2005, LSOK2011, MSUP2007}. Mass balance alone depends on being able to accurately estimate the volume proportions of solid phases and quench phases, recognise and subtract quench overgrowths where they exist, and calculate the effective ambient-pressure density of the melt. All of these are associated with significant uncertainties. Deficits in microprobe totals are sometimes reasonable for estimating water contents in solid phases, provided the microprobe standards are suitable, and the microprobe is exceptionally well-calibrated (even a small deficit implies a large water content). However, for quench melts the deficits are attributable to water held in quench phases \emph{and} porosity within the volume sampled by the defocused beam. For this reason, defocused beam techniques should probably be restricted to estimating non-volatile element ratios (such as Mg:Si).

@@

These equations can be used to describe in an average sense a range of melt-melt equilibria involving monomers (Mg(OH)$_2$ and Si(OH)$_4$), dimers (Mg$_2$O(OH)$_2$, MgSiO(OH)$_4$, Si$_2$O(OH)$_6$) and higher oligomers. A high proportion of monomers and dimers exist in relatively dilute solutions \citep{XXXX}, but not in the concentrated solutions investigated in this study.

@@

These models can be used to calculate solid-melt equilibria by equating the Gibbs free energy of the solid $G^s_{solid}$ with the chemical potential of the melt component with the same composition $\mu^s_{melt}$:
\begin{equation}
G^s_{solid} = \mu^s_{melt} = G^s_{melt} - RT \ln a^s_{melt}
\label{eqn:equilibrium}
\end{equation}
In this study, we use the thermodynamic dataset of \cite{SLB2011} to calculate $G^s_{solid}$ for periclase, forsterite and its high pressure polymorphs, orthoenstatite, high pressure clinoenstatite, coesite and stishovite. A modified version of the molecular dynamics-derived anhydrous melt model of \cite{DKS2013} is used to calculate $G^s_{melt}$, and the \cite{SS1985} model is used to calculate $a^s_{melt}$. The calculations in the work were undertaken with the mineral physics toolkit Burnman \citep[available from \url{https://geodynamics.org/cig/software/burnman/}][]{CHRU2014}.

\subsection{The anhydrous melt model}
\subsubsection{Endmembers}
Thermodynamic models of MgO-SiO$_2$ melts are currently in their infancy. Many of the techniques used to constrain the thermodynamic properties (particularly the equations of state) of solids cannot be used on liquids as a result of their highly dynamic behaviour. Therefore in this study, we use MgO and SiO$_2$ melt models derived from molecular dynamics simulations \citep{DKS2013}, and fit a simplified mixing curve based on experimental data. First, however, the thermodynamic properties of the endmember melts need to be adjusted to fit the experimentally derived thermodynamic data at the melting point. This step was unnecessary in the original study of \cite{DKS2013} because solid properties were also obtained by molecular dynamics simulations. Such simulations do not provide good estimates of absolute energies and entropies, and so the values of the solids were obtained by fixing the melting point at a given pressure to the experimentally derived values (0 GPa, 3070 K for periclase, 14 GPa, 3120 K for stishovite). It turns out that the MD-determined entropy of periclase is almost identical to the experimentally determined values at all pressures and temperatures \citep{SLB2011}, so we only needed to adjust the standard state free energy. This leads to a periclase melting point of $\sim$ 4180 K at 13 GPa, similar to that determined by previous ab-initio studies \citep{Alfe2005}. We note that experimental results vary by about 1000 K about this value \citep[3100 -- 5373 K at 13 GPa][]{ZB1994,ZF2008}. Current thermodynamic models for periclase cannot be extrapolated to the extreme temperatures predicted by the \cite{ZF2008} data. However, assuming that the volume change of melting is roughly constant with temperature (1.8 cm$^3$/mol at 13 GPa), the entropy of melting at a melting point of 5373 K would be about 50\% of that for a melting point of 4180 K (13.25 vs. 27.5 J/K/mol). At 13 GPa and 2000 K, MgO melt would actually be 15 kJ/mol \emph{more} stable as a result of this lower entropy.

The MD-determined entropy of stishovite is significantly lower than the experimentally determined value \citep{SLB2011}, so we increased the standard state entropy of the melt by 16.5 J/K/mol. Conveniently, we also have a way to test the volumes predicted by the model of \cite{DKS2013} at around 13 GPa. The experimentally-derived slope of the coesite melting curve is essentially zero at the triple point with stishovite at 13.7 GPa \citep{ZLGHF1993}. Therefore, from the Clausius-Clapeyron relationship $dT/dP = \Delta V / \Delta S$ the volumes of coesite and SiO$_2$ melt must be zero at this point. Unfortunately, the MD-derived SiO$_2$ melt model has its density crossover with coesite about 4 GPa lower. Therefore, we adjusted the bulk modulus and reference pressure to satisfy this constraint, and the volume at zero pressure. The adjustments required here may not be a failing of the MD simulations; it is likely that the coordination changes in the SiO$_2$ melt require a higher order Birch Murnaghan EoS than that used by \cite{DKS2013}. Our adjusted model is probably only accurate in a range of about $\pm$5 GPa around the triple point. The experimental data can also be used to judge the accuracy of the entropy of the SiO$_2$ melt model. The MD-derived entropy of melting is $\sim$40 J/K/mol at the triple point. This value is in perfect agreement with the value derived from the Clapeyron slope of the stishovite liquidus \citep{SL1995}. We note that the slope recently proposed by \cite{Millotetal2015} is much shallower, but their slope is a two parameter Simon's equation fit based on the triple point temperature at 13.7 GPa and the melting point at 500 GPa, and fits the data at 15--100 GPa \citep{SL1995, LAM1983} rather poorly. A Simon's equation fit is probably inappropriate over such extreme pressure ranges, especially where the melting point may be rather sensitive to changes in cation coordination.

\subsubsection{Solution model}
A subregular solution model was adopted to describe the free energy of intermediate-composition MgO-SiO$_2$ melts. The two interaction parameters W$_{MS}$ and W$_{SM}$ at 13 GPa were derived from the melt composition at the forsterite (fo)-HP clinoenstatite (hen) eutectic \citep{PWMW1998} and melting temperatures at the fo, hen and fo-hen cotectic compositions \citep{PG1990, PW1993}. We assumed an ideal entropic contribution to mixing. The pressure dependencies were fit to the melting curves. It should be noted that although the excess volumes are similar to the results of \cite{DKS2013}, the excess entropies are some 50\% lower. However, they provide an excellent fit to the forsterite and orthoenstatite melting curves (see Supplementary Materials).

\subsection{The hydrous melt model}

@@

\begin{equation}
p' = \frac{p_{H_2O}}{p_{H_2O}  + r \left(1.-p_{H_2O} \right)} 
\end{equation}
\begin{eqnarray}
X_O = 1 - p' - \frac{ 0.5 - \sqrt{ 0.25 - \frac{K_{(1)}-4}{K_{(1)}} \left(p'-p'^2 \right)}}{\frac{K_{(1)}-4}{K_{(1)}} }\\
X_{H_2O} = X_O + 2 p' - 1 \\
X_{OH} = 1 - X_O - X_{H_2O}
\end{eqnarray}
\noindent $p'$ is the molar proportion of H$_2$O where the silicate is normalised on a one oxygen basis (rather than a one-cation basis). We assume that all bridging oxygens are equally likely to be protonated. The number of oxygens per formula unit $r$ thus increases monotonously from 1 to 2 across the MgO-SiO$_2$ binary ($r_{\textrm{per}}$ = 1, $r_{\textrm{fo, wad, rw}}$ = 4/3, $r_{\textrm{oen, cen}}$ = 3/2, and $r_{\textrm{coe, stv}}$ = 2). The ideal activities of the anhydrous and water components can then be calculated:
\begin{eqnarray}
a^{\textit{s}}_{\textit{ideal melt}} =X_O^r \\
a^{H_2O}_{\textit{ideal melt}} = X_{H_2O}
\end{eqnarray}

The activity coefficients (i.e. the activity prefactor which provides the non-ideal contribution to mixing; $a = \gamma a_{ideal}$) are equal to one except in the case of the MgO-H$_2$O binary, where a simple Margules model is applied \citep{HM2012}
\begin{equation}
RT \ln \gamma_i = {p_j}^2 \left( 1 - p_j \right) \left( 2 W_{ji} - W_{ij} \right) \\
\end{equation}
\noindent where the subscripts $i$ and $j$ correspond to the components $s$ or H$_2$O. Equilibrium with the solid can now be found using Equation \ref{eqn:equilibrium}, noting that the dataset-derived values of $G^s_{solid}$ of the forsterite (Mg$_2$SiO$_4$) and enstatite (Mg$_2$Si$_2$O$_6$) polymorphs should be divided by 3 and 4 to reduce them to their one-cation equivalents.

@@

\clearpage
\subsubsection{Mg$_2$SiO$_4$-H$_2$O}
The melting point depression of forsterite in the presence of water at 13 GPa is illustrated in Figure \ref{fig:foH}. The model parameters shown are $G_{(1)} = 75 \cdot (1200-T)$ J/mol, $r=4/3$. The anhydrous melting of forsterite is incongruent, so at low water contents the liquidus phase along the Mg$_2$SiO$_4$-H$_2$O binary is periclase (see also Figure \ref{fig:ternary}). Water-rich chambers from experiments run at 1300 $^{\circ}$C revealed small quantities of enstatite, indicating that the forsterite-enstatite cotectic crosses the Mg$_2$SiO$_4$-H$_2$O tie line at 1300-1400 $^{\circ}$C. This is in excellent agreement with results on the wadsleyite-water system run at 15 GPa \citep{DDFK2005, LSOK2011}. At 1200 $^{\circ}$C, one chamber contained large crystals of hydroxychondrodite. The presence of this MgO-rich phase in the chamber with the lowest water contents may be explained by an MgO-deficit in forsterite resulting from the incorporation of hydrogen via the substitution mechanism Mg$^{2+}$ $\rightarrow$ 2H$^+$ \citep{KB2006}. Alternatively it could simply be the result of a small SiO$_2$ deficit in that sample chamber.

@@

\subsubsection{MgSiO$_3$-H$_2$O}
The melting of high pressure clinoenstatite in the presence of water is shown in Figure \ref{fig:eoH}. Unlike the forsterite or periclase liquidi, the observed depression of melting at low water contents exceeds that predicted by even the most extreme commonly-used ideal mixing model, where all oxygens in the melt are equivalent and available for protonation, and molecular H$_2$O is absent \citep{SS1985}. The liquidus at higher water contents can be fit well with model parameters $G_{(1)} = 120 \cdot (1750-T)$ J/mol, $r=3/2$. The disagreement between this model and the experimental observations is small, and could plausibly be due to experimental error. However, given the results for stishovite (see next section), we feel that more complex models are required to explain the depression of melting in the MgSiO$_3$ system.

@@

Our experimentally determined liquidus is in excellent agreement with the experimental results of \cite{YII2004}, as is the absence of stishovite in any of the sample chambers. We note that although we do not observe any evidence for melt-fluid immiscibility, the flat liquidus at ca. 1500 $^{\circ}$C causes a very rapid change from water-rich to water-poor compositions with increasing temperature, in agreement with \cite{Inoue1994} and \cite{YII2004}. One possibility is that this reflects relatively high water loss from the water rich bulk compositions. However, there is very little evidence for melt loss from the capsules; the lids on the capsules were well-sealed and showed no evidence of melt escape. The implication is that for a given bulk H$_2$O content, a sharp rise in the degree of partial melting occurs when bulk compositions along the en+H$_2$O join are heated above 1500 $^{\circ}$C.

\clearpage
\subsubsection{SiO$_2$-H$_2$O}
The 13 GPa SiO$_2$-H$_2$O binary phase diagram is shown in Figure \ref{fig:SH}. While the deviation between the predicted enstatite liquidus depression for ideal molecular H$_2$O-free melts and the experimental observations are relatively small, the deviation for the stishovite data is extreme. The mixing model presented in Figure \ref{fig:SH} has parameters $G_{(1)} = 100 \cdot (1200-T)$ J/mol, $r=2$ is fit to the extrapolation of the experimentally defined liquiudus, but clearly fails to fit any of the experimental data. There are several possible explanations for this. First, the melting point of anhydrous stishovite (which is metastable with respect to coesite) could be overestimated by ca. 300 $^{\circ}$C, or the entropy of melting could be 10--20 J/K/mol, rather than 40 J/K/mol. Neither of these possibilities is very likely; the coesite and stishovite melting curves provide good constraints on the volume and entropy of SiO$_2$ melt in the region of the triple point at ca. 13.5 GPa. A third possibility is that liquidus-phase stishovite crystals were missed during SEM analysis. In the most extreme cases, the missing fraction of crystals would have been on the order of 30\%. This is also extremely unlikely; the equant and dendritic morphologies of liquidus and quench stishovite are instantly recognisable, and the thermal gradients in the capsule lead to melt separation on the timescales of the experiments, such that even a small (5\%) volume proportion of solids should be detected.

@@

Here, we recalculate the wadsleyite/melt and ringwoodite/melt partitioning values with our melt model, assuming that our fitted values of $G_{(1)}$ and $r$ are suitable between 13 and 20 GPa. We justify this by noting that speciation is likely to be only weakly dependent on pressure under these conditions. Water concentrations in the solid are taken from the original studies. We restrict our analysis to data from 1100--1400 $^{\circ}$C, where the Mg:Si ratios of the melts are similar to that in forsterite. At 15 GPa, $D^{wad/melt}_{H_2O}$ increases from 0.30 to 0.51 between 900 and 1200 $^{\circ}$C, and then decreases back to 0.32 at 1400 $^{\circ}$C. The three ringwoodite H$_2$O concentrations (20-23 GPa) indicate a decrease in $D^{rw/melt}_{H_2O}$ from 0.68 to 0.57 between 1300 and ca. 1400 $^{\circ}$C.

@@

% Future work
We hope that this work forms the foundation for more detailed thermodynamic models of hydrous mantle melts. Particularly intriguing are the concave up topologies of the forsterite, enstatite and stishovite liquidus surfaces, and the large depression of the melting point at low water contents in silica-rich systems. Both of these observations imply efficient hydroxylation of the silicate network and strong interactions between species, which would be interesting subjects for molecular dynamic studies. We also hope that a combination of our liquidus constraints with parameterisations for density and viscosity will be used to inform more realistic models of melt migration above the mantle transition zone and during subduction.

\section{Acknowledgements}
The authors would like to thank Gerti Gollner for acquiring capsule materials, Stefan \"Ubelhack and Heinz Fischer for assembly and capsule cutting and Hubert Schulze for sample preparation. RM was supported by a Humboldt Postdoctoral Fellowship. This study was funded by the ACCRETE project (European Research Council Advanced Grant, contract number 290568).

>>>>

%% Group authors per affiliation:
\author{R. Myhill}
\address{Bayerisches Geoinstitut, Universit\"{a}t Bayreuth, Universit\"{a}tsstrasse 30, 95447 Bayreuth, Germany \\ Now at School of Earth Sciences, University of Bristol, Wills Memorial Building, Queens Road, Bristol BS8 1RJ, United Kingdom}
\cortext[mycorrespondingauthor]{Corresponding author: R. Myhill}
\ead{myhill.bob@gmail.com}

@@

\section{Introduction}
Hydrous melts have a remarkable influence on Earth's surface and interior. Melting at subduction zones is primarily driven by the release of hydrous fluids from the warming slab at about 100 km depth, producing the world's arc volcanoes. The water cycle does not stop here, however; hydrous minerals within the crust should allow some water to be transported to 250 km \citep{PS2002}, and peridotite hydration should allow water to reach the mantle transition zone or deeper, providing temperatures remain sufficiently low \citep{KOM2005}. This deep water cycle must extend to at least 500 km depth, as implied by the discovery of water-saturated ringwoodite in a diamond from Juina, Brazil \citep{Pearsonetal2014}. In addition to kimberlites and arc volcanoes, hydrous melts are known to metasomatise and refertilise the lithospheric mantle, creating the so-called MARID (mica, amphibole, rutile, ilmenite, diopside) assemblage \citep{GHKR2010}. It has also been argued that neutral or negatively buoyant water-rich melts are responsible for low velocity and high conductivity layers observed at the 410 km discontinuity, and that these small fraction melts could act to filter out incompatible elements during upwelling \citep{BK2003}.

@@

Melt compositions are also required to understand the stability of dense hydrous magnesium silicate (DHMS) phases and the water capacity of nominally anhydrous minerals (NAMS) at high pressure. Together, these control water transport and storage in the deep mantle, and as such are the primary inputs to models of the water budget of the Earth \citep{Hirschmann2006}. Hydrous fluids at elevated pressures and temperatures cannot be treated as pure H$_2$O. As pressure increases, immiscibility between fluid and melt decreases \citep{MSUP2007, SUTG2001, HM2012}, and the thermal stability of Mg-rich hydrous phases increases \citep{KO2006}. As a result, the high pressure breakdown of hydrous phases releases fluids which have H$_2$O activities significantly lower than one. Understanding and predicting the \emph{P-T} conditions of DHMS phase breakdown and the water storage capacity at of NAMS therefore requires good activity-composition models for hydrous melts.

Constraining melt compositions remains a major challenge for the field of high-pressure experimental petrology. The primary problem is that water-rich melts are unquenchable. During quenching, some of the water in the melt crystallises as hydrous minerals such as brucite, but a significant proportion crystallises as water ice at high pressure, which is then lost as water vapour during decompression and subsequent sample preparation. The original H$_2$O contents of the partial hydrous melts are normally estimated by mass balance and/or by deficits in microprobe totals \citep[e.g.][]{OMY2000, DDFK2005, LSOK2011, MSUP2007}. Using mass balance to calculate H$_2$O contents depends on being able to estimate the volume proportions of solid and quench phases, recognise and subtract quench overgrowths where they exist, and calculate the effective ambient-pressure density of the melt. Such estimates are typically associated with significant uncertainties. Deficits in microprobe totals are sometimes reasonable for estimating water contents in solid phases, provided the microprobe standards are suitable, and the microprobe is exceptionally well-calibrated (even a small deficit implies a large water content). However, for quench melts the deficits are attributable to water held in quench phases \emph{and} porosity within the volume sampled by the defocused beam. For this reason, defocused beam techniques should probably be restricted to estimating non-volatile element ratios (such as Mg:Si).

@@

These equations can be used to describe in an average sense a range of melt-melt equilibria involving monomers (Mg(OH)$_2$ and Si(OH)$_4$), dimers (Mg$_2$O(OH)$_2$, MgSiO(OH)$_4$, Si$_2$O(OH)$_6$) and higher oligomers. Concentrated solutions such as those investigated in this study are likely to be dominated by highly polymerised oligomers \citep[see][]{HM2012}.

@@

These models can be used to calculate solid-melt equilibria by equating the Gibbs free energy of the solid $G^s_{solid}$ with the chemical potential of the melt component with the same composition $\mu^s_{melt}$:
\begin{equation}
G^s_{solid} = \mu^s_{melt} = G^s_{melt} - RT \ln a^s_{melt}
\label{eqn:equilibrium}
\end{equation}
In this study, we use the thermodynamic dataset of \cite{SLB2011} to calculate $G^s_{solid}$ for periclase, forsterite and its high pressure polymorphs, orthoenstatite, high pressure clinoenstatite, coesite and stishovite. A modified version of the molecular dynamics-derived anhydrous melt model of \cite{DKS2013} is used to calculate $G^s_{melt}$, and the \cite{SS1985} model is used to calculate $a^s_{melt}$. The calculations in the work were undertaken with the mineral physics toolkit Burnman \citep[available from \url{https://geodynamics.org/cig/software/burnman/};][]{CHRU2014}.

\subsection{The hydrous melt model}
\subsubsection{Anhydrous melt endmembers}
Thermodynamic models of MgO-SiO$_2$ melts are currently in their infancy. Many of the experimental techniques used to constrain the thermodynamic properties (particularly the equations of state) of solids cannot be used on liquids as a result of their highly dynamic behaviour. In this study, we instead use MgO, Mg$_2$SiO$_4$, MgSiO$_3$ and SiO$_2$ melt models derived from molecular dynamics simulations \citep{DKS2013}. First, however, the thermodynamic properties of the endmember melts need to be adjusted to fit the experimentally derived thermodynamic data at the melting point. This step was unnecessary in the original study of \cite{DKS2013} because solid properties were also obtained by molecular dynamics simulations. Such simulations do not provide good estimates of absolute energies and entropies, and so the values of the solids were obtained by fixing the melting point at a given pressure to the experimentally derived values (0 GPa, 3070 K for periclase, 14 GPa, 3120 K for stishovite). It turns out that the MD-determined entropy of periclase is almost identical to the experimentally determined values at all pressures and temperatures \citep{SLB2011}, so we only needed to adjust the standard state free energy. This leads to a periclase melting point of $\sim$ 4180 K at 13 GPa, similar to that determined by previous ab-initio studies \citep{Alfe2005}. We note that experimental results vary by about 1000 K about this value \citep[3100 -- 5373 K at 13 GPa;][]{ZB1994,ZF2008}. It is therefore worth considering the effect of different melting curves on the free energy difference between melt and solid. Assuming that the volume change of melting is roughly constant with temperature (1.8 cm$^3$/mol at 13 GPa), the entropy of melting at a melting point of 5373 K would be about 50\% of that for a melting point of 4180 K (13.25 vs. 27.5 J/K/mol). The difference in free energy would therefore increase more slowly as temperature decreased, reducing the discrepancy between the two models at relatively low temperatures.

The treatment for Mg$_2$SiO$_4$ and MgSiO$_3$ was similar to that for periclase; the standard state energies of the liquids were adjusted to fit the melting points at 13 GPa for forsterite \citep[2290 $^{\circ}$C][]{PW1993} and HP-clinoenstatite \citep[2240 $^{\circ}$C][]{PG1990}, using the endmember models of \cite{SLB2011}.

In the case of SiO$_2$ melt, we used the experimentally-derived melting curve to assess the \cite{DKS2013} model. The slope of the coesite melting curve is essentially zero at the triple point with stishovite at 13.7 GPa \citep{ZLGHF1993}, such that $V$ (coesite) = $V$ (SiO$_2$ melt) at this point. The slope of the stishovite melting curve at the triple junction \citep{SL1995} then provides an estimate of the entropy of melting (via the Clausius-Clapeyron relationship $dT/dP = \Delta V / \Delta S$), which is $\sim$40 J/K/mol. We note that the stishovite melting curve recently proposed by \cite{Millotetal2015} is much shallower than that constrained by experiments at 15--100 GPa \citep{SL1995, LAM1983}. Their two-parameter Simon's equation fit between the melting points at 13.7 and 500 GPa is probably inappropriate over such extreme pressure ranges, especially where the melting point may be rather sensitive to changes in cation coordination.

At the triple point, the MD-derived entropy of melting is in excellent agreement with the experimental value, but the entropy of stishovite is 16.4 J/K/mol lower than the experimentally determined value \citep{SLB2011}. For this reason, we increased the standard state entropy of the SiO$_2$ melt by 16.4 J/K/mol. We also adjusted the bulk modulus and reference pressure to obtain the correct pressure at the coesite-melt density crossover. The adjustments required here may not be a failing of the MD simulations; the coordination changes in the SiO$_2$ melt may require a higher order EoS than that used by \cite{DKS2013}. Our adjusted model is probably only accurate in a range of about $\pm$5 GPa around the triple point.

\subsubsection{The hydrous melt model}

@@

\begin{equation}
p' = \frac{p_{H_2O}}{p_{H_2O}  + r \left(1.-p_{H_2O} \right)} 
\end{equation}
\begin{eqnarray}
X_O = 1 - p' - \frac{ 0.5 - \sqrt{ 0.25 - \frac{K_{(1)}-4}{K_{(1)}} \left(p'-p'^2 \right)}}{\frac{K_{(1)}-4}{K_{(1)}} }\\
X_{H_2O} = X_O + 2 p' - 1 \\
X_{OH} = 1 - X_O - X_{H_2O}
\end{eqnarray}
\noindent $p'$ is the molar proportion of H$_2$O where the silicate is normalised on a one oxygen basis (rather than a one-cation basis). We assume that all bridging oxygens are equally likely to be protonated. The number of oxygens per formula unit $r$ thus increases monotonously from 1 to 2 across the MgO-SiO$_2$ binary ($r_{MgO}$ = 1, $r_{Mg_2SiO_4}$ = 4/3, $r_{MgSiO_3}$ = 3/2, and $r_{SiO_2}$ = 2). The ideal activities of the anhydrous and water components can then be calculated:
\begin{eqnarray}
a^{\textit{s}}_{\textit{ideal melt}} =X_O^r \\
a^{H_2O}_{\textit{ideal melt}} = X_{H_2O}
\end{eqnarray}

The activity coefficients (i.e. the activity prefactor which provides the non-ideal contribution to mixing; $a = \gamma a_{ideal}$) are equal to one except in the case of the MgO-H$_2$O binary, where a simple Margules model is applied \citep{HM2012}
\begin{equation}
RT \ln \gamma_i = {p_j}^2 \left( 1 - p_j \right) \left( 2 W_{ji} - W_{ij} \right) \\
\end{equation}
\noindent where the subscripts $i$ and $j$ correspond to the components $s$ or H$_2$O. Equilibrium with the solid can now be found using Equation \ref{eqn:equilibrium}, noting that the dataset-derived values of $G^s_{solid}$ of the forsterite (Mg$_2$SiO$_4$) and enstatite (Mg$_2$Si$_2$O$_6$) polymorphs \citep{SLB2011} should be divided by 3 and 4 to reduce them to their one-cation equivalents. Similarly, the values of $G^s_{melt}$ for the MD-derived Mg$_2$SiO$_4$ and MgSiO$_3$ models \citep{DKS2013} should be divided by 3 and 2 respectively.

@@

\clearpage
\subsubsection{Mg$_2$SiO$_4$-H$_2$O}
The melting point depression of forsterite in the presence of water at 13 GPa is illustrated in Figure \ref{fig:foH}. The model parameters shown are $G_{(1)} = 75 \cdot (1420-T)$ J/mol, $r=4/3$. The anhydrous melting of forsterite is incongruent, so at low water contents the liquidus phase along the Mg$_2$SiO$_4$-H$_2$O binary is periclase (see also Figure \ref{fig:ternary}). Water-rich chambers from experiments run at 1300 $^{\circ}$C revealed small quantities of enstatite, indicating that the forsterite-enstatite cotectic crosses the Mg$_2$SiO$_4$-H$_2$O tie line at 1300-1400 $^{\circ}$C. This is in excellent agreement with results on the wadsleyite-water system run at 15 GPa \citep{DDFK2005, LSOK2011}. At 1200 $^{\circ}$C, one chamber contained large crystals of hydroxychondrodite. The presence of this MgO-rich phase in the chamber with the lowest water contents may be explained by an MgO-deficit in forsterite resulting from the incorporation of hydrogen via the substitution mechanism Mg$^{2+}$ $\rightarrow$ 2H$^+$ \citep{KB2006}. Alternatively it could simply be the result of a small SiO$_2$ deficit in that sample chamber.

@@

\subsubsection{MgSiO$_3$-H$_2$O}
The melting of high pressure clinoenstatite in the presence of water is shown in Figure \ref{fig:eoH}. Unlike the forsterite or periclase liquidi, the observed depression of melting at low water contents exceeds that predicted by even the most extreme commonly-used ideal mixing model, where all oxygens in the melt are equivalent and available for protonation, and molecular H$_2$O is absent \citep{SS1985}. The liquidus at higher water contents can be fit well with model parameters $G_{(1)} = 190 \cdot (1700-T)$ J/mol, $r=3/2$. The disagreement between this model and the experimental observations is small, and could plausibly be due to experimental error. However, given the results for stishovite (see next section), we feel that MgSiO$_3$-H$_2$O melts exhibit significant nonideality even at 13 GPa.

@@

Our experimentally determined liquidus is in excellent agreement with the experimental results of \cite{YII2004}, as is the absence of stishovite in any of the sample chambers. We note that although we do not observe any evidence for melt-fluid immiscibility, the flat liquidus at ca. 1500 $^{\circ}$C causes a very rapid change from water-rich to water-poor compositions with increasing temperature, in agreement with \cite{Inoue1994} and \cite{YII2004}. One possibility is that this reflects relatively high water loss from the water rich bulk compositions. However, there is very little evidence for melt loss from the capsules; the lids on the capsules were well-sealed and showed no evidence of melt escape. The implication is that for a given bulk H$_2$O content at 13 GPa, a sharp rise in the degree of partial melting occurs when bulk compositions along the en+H$_2$O join are heated above 1500 $^{\circ}$C.

\clearpage
\subsubsection{SiO$_2$-H$_2$O}
The 13 GPa SiO$_2$-H$_2$O binary phase diagram is shown in Figure \ref{fig:SH}. While the deviation of the experimental results from the predicted enstatite liquidus depression for ideal molecular H$_2$O-free melts was relatively small, the deviation for the stishovite results is extreme. The mixing model presented in Figure \ref{fig:SH} has parameters $G_{(1)} = 100 \cdot (1200-T)$ J/mol, $r=2$ is fit to the extrapolation of the experimentally defined liquidus, but clearly fails to fit any of the experimental data. There are several possible explanations for this. The melting point of anhydrous stishovite (which is metastable with respect to coesite at 13 GPa) could be overestimated by ca. 300 $^{\circ}$C, or the entropy of melting could be 10--20 J/K/mol, rather than 40 J/K/mol. Neither of these possibilities seems very likely; the coesite and stishovite melting curves \citep{ZLGHF1993} provide good constraints on the volume and entropy of SiO$_2$ melt in the region of the triple point at ca. 13.5 GPa. An alternative possibility is that liquidus-phase stishovite crystals were missed during SEM analysis. In the most extreme cases, the missing fraction of crystals would have to have been on the order of 30\%. This is also extremely unlikely; the equant and dendritic morphologies of liquidus and quench stishovite are instantly recognisable, and the thermal gradients in the capsule lead to melt separation on the timescales of the experiments, such that even a small (5\%) volume proportion of solids should be detected.

@@

Here, we recalculate the wadsleyite/melt and ringwoodite/melt partitioning values with our melt model, assuming that our fitted values of $G_{(1)}$ and $r$ are suitable between 13 and 20 GPa. We justify this by noting that speciation is likely to be only weakly dependent on pressure under these conditions. We also approximate the activity of Mg$_2$SiO$_4$ in the solid phases to be 1, such that the additional water in wadsleyite and ringwoodite does not affect the calculated liquidus curves. Water concentrations in the solid are taken from the original studies. We restrict our analysis to data from 1100--1400 $^{\circ}$C, where the Mg:Si ratios of the melts are similar to that in forsterite. At 15 GPa, $D^{wad/melt}_{H_2O}$ increases from 0.30 to 0.51 between 900 and 1200 $^{\circ}$C, and then decreases back to 0.32 at 1400 $^{\circ}$C. The three ringwoodite H$_2$O concentrations (20-23 GPa) indicate a decrease in $D^{rw/melt}_{H_2O}$ from 0.68 to 0.57 between 1300 and ca. 1400 $^{\circ}$C.

@@

% Future work
We hope that this work forms the foundation for more detailed thermodynamic models of hydrous mantle melts. Particularly intriguing are the concave up topologies of the forsterite, enstatite and stishovite liquidus surfaces, and the large depression of the melting point at low water contents in silica-rich systems. Both of these observations imply efficient hydroxylation of the silicate network and strong interactions between species, which would be interesting subjects for molecular dynamic studies. We also hope that a combination of our liquidus constraints with parameterisations for density and viscosity will be used to inform more realistic models of hydrous melt migration in the deep mantle.

\section{Acknowledgements}
The authors would like to thank Gerti Gollner for acquiring capsule materials, Stefan \"Ubelhack and Heinz Fischer for assembly and capsule cutting and Hubert Schulze for sample preparation. RM was supported by a Humboldt Postdoctoral Fellowship. This study was funded by the ACCRETE project (European Research Council Advanced Grant, contract number 290568). Python files recreating the results from this paper may be downloaded from the BurnMan webpage (\url{https://geodynamics.org/cig/software/burnman/}), and are also available from RM.
